Differential abundance analysis of taxonomically-biased microbiome measurements

### 2022-03-09 appendix-regression.Rmd
Add gamma-Poisson regression to appendix
Introduces gamma-Poisson regression as it applies to sequencing data,
and describes how to set the offsets to correct for bias when estimated
efficiencies are available from community control measurements.

# Taxonomic bias and linear regression {#appendix-regression}

@@

## Review of simple linear regression

@@

## Measurement error in the response

@@

## Specific application to taxonomic bias

@@

>>>>

# Linear regression {#appendix-regression}

## Simple linear regression

@@

### Review of simple linear regression

@@

### Measurement error in the response

@@

### Specific application to taxonomic bias

@@

## Gamma-Poisson regression

This section describes how gamma-Poisson regression can, with appropriate choice of offsets, be used to estimate LFCs in proportions from MGS count data either with or without bias correction.

### Background

The gamma-Poisson distribution, also known as the negative binomial distribution, is a distribution that is commonly used to model sequencing count data (@holmes2018mode).
Its key advantage over the simple linear regression model is that it directly models the random sampling of reads that occurs during a sequencing experiment and, in this way, naturally accounts for the imprecision associated with an observation of zero or a small positive count for a given species and sample.
The gamma-Poisson distribution has two parameters, which jointly determine the mean and the standard deviation.
We use the parameterization used by the rstanarm R package (@rstanarm), which we use for fitting gamma-Poisson GLMs in our case-study analyses.
Suppose that $y^{(a)}$ has a gamma-Poisson distribution conditional on a covariate $x$.
Define two parameters $\mu$ and $\phi$ such that
\begin{align}
  E[y^{(a)} \mid x]  &= \mu^{(a)} \\
  sd[y^{(a)} \mid x] &= \sqrt{
    E[y^{(a)} \mid x] + E[y^{(a)} \mid x]^2 / \phi
  };
\end{align}
thus $\mu$ equals the mean and $\phi$, known as the 'reciprocal dispersion', increases the standard deviation as it approaches 0 and causes the standard deviation to approach $\sqrt{\mu}$ (that of a Poisson distribution) as it approaches infinity.

For use in gamma-Poisson GLM regression of MGS data, we partition the mean into two factors, $\mu^{(a)} = u^{(a)} \theta^{(a)}$ (@gelman2020regr).
Here, $\theta^{(a)} = e^{\alpha + \beta x^{(a)}}$ is the quantity of primary interest, whose average LFC equals $\beta$.
The factor $u^{(a)}$ is a sample-specific _exposure_ parameter that determines the overall scale of counts.
We can equivalently write $\mu^{(a)} = e^{\alpha + \beta x^{(a)} + \log u^{(a)}}$;
the logarithm of the exposure, $\log u^{(a)}$, is known as the _offset_ of the GLM.
Performing gamma-Poisson regression with standard statistical software requires that the offsets are provided as known values.

### Inferring LFCs in proportions with and without bias correction

Our goal is to use gamma-Poisson regression to estimate the LFC in a species proportion from the observed read counts. 
We therefore equate the counts $M_i^{(a)}$ with $y^{(a)}$ and the proportion $P_{i}^{(a)}$ with $\theta^{(a)}$; it remains to determine the offsets that are consistent with our MGS model.
To do so, we relax the deterministic assumption of our original model and instead suppose that the right-hand side of Equation \@ref(eq:mgs-model) equals the _expected_ read count of species $i$ in sample $a$,
\begin{align}
  (\#eq:mgs-model-gp)
  E[M_i^{(a)}] = A_i^{(a)} B_i F^{(a)}.
\end{align}
The total expected count is
\begin{align}
  (\#eq:total-reads-gp)
  E[M_S^{(a)}] &= \sum_{i \in S} E[M_i^{(a)}]
  \\&= \left( \sum_{i \in S} A_i^{(a)} B_i \right) F^{(a)} 
  \\&= A_S^{(a)} B_S^{(a)} F^{(a)}.
\end{align}
Under the gamma-Poisson model for $M_i^{(a)}$, we have that $E[M_i^{(a)}] = u_i^{(a)} P_i^{(a)}$.
Equating this expression for $E[M_i^{(a)}]$ with Equation \@ref(eq:mgs-model-gp) lets us solve for the exposure,
\begin{align}
  (\#eq:exposure)
  u_i^{(a)} 
    &= \frac{A_i^{(a)} B_i F^{(a)}}{P_i^{(a)}} 
  \\&= B_i A_S^{(a)} F^{(a)}
  \\&= \frac{B_i}{B_S} \cdot E[M_S^{(a)}].
\end{align}
The second line follows from $P_i^{(a)} \equiv A_i^{(a)}/ A_S^{(a)}$, and the third line from Equation \@ref(eq:total-reads-gp).

The final expression in Equation \@ref(eq:exposure) can be used to compute offsets for the regression.
The three terms $B_i$, $B_S^{(a)}$, and $E[M_S^{(a)}]$ are each unknown, so we instead substitute estimates for these terms to obtain an estimated exposure $\hat u_{i}^{(a)}$; we then set the offset in the linear model to $\log \hat u_{i}^{(a)}$.
We can estimate $E[M_S^{(a)}]$ by the observed total count, $M_S^{(a)}$.
In the absence of bias correction, we set $B_{i} = B_{S}^{(a)} = 1$; our estimate of the exposure is then just
\begin{align}
  \hat u_i^{(a)} = M_S^{(a)}.
\end{align}
We can apply bias correction using estimates of the efficiencies, $\hat B_{i}$, derived from community control measurements.
From these estimates and the observed counts, we can estimate the mean efficiency in the sample by
\begin{align}
  \hat B_S^{(a)} = \sum_{i\in S} B_i \hat P_i^{(a)},
\end{align}
where $\hat P_i^{(a)}$ are the calibrated proportions obtained from a simple plug-in procedure,
\begin{align}
  \hat P_i^{(a)} = \frac{M_i^{(a)} / B_i}{\sum_j M_j^{(a)} / B_j}.
\end{align}
Equivalently, we can calculate $\hat B_S^{(a)}$ directly from the observed counts,
\begin{align}
  \text{B}_S^{(a)} =
  \left[\sum_j \frac{M_j^{(a)}}{M_S^{(a)} \text{B}_j}\right]^{-1}.
\end{align}
We then estimate the exposure by
\begin{align}
  \hat u_i^{(a)} = \frac{\hat B_i}{\hat B_S} \cdot M_S^{(a)}.
\end{align}
